Equity conceptualizations and standards within transportation literature: a scoping review

### 2023-10-03 Manuscript/Manuscript.qmd
More edits to Manuscript.qmd

Equitable transportation system from a planning perspective, involves fair production, distribution, and reproduction of transportation benefits and costs within a community [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021].
Transportation equity is a multi-scalar and multi-dimensional topic centered on the distribution of transport-related benefits and burdens across population groups and communities within urban, suburban or rural areas.
Benefits and burdens can include, but are not limited to, accessibility, mobility, affordability, health and environmental issues that impact individuals, marginalized members of a community, and the community as a whole [@lucasMeasuringTransportEquity2019].

Transportation systems are complex: they support multi-modal uses and are home to emerging new technologies and service modes that complicate the balance of benefit and burdens [@guoSystematicOverviewTransportation2020].
Consequently, achieving an *equitable* system is not a positivist attribute; it should be normatively defined by the community  and broader society [@pereiraTransportationEquity2021; @paezMeasuringAccessibilityPositive2012] during the ongoing process of measurement, operationalization, and improvement.
Both academic literature and planners are grappling to contextualize what an equitable transportation system may look like along spatial, temporal, environmental, and socio-economical dimensions [@boisjolyHowGetThere2017; @pereiraTransportationEquity2021].
Synthesizing methods on how to identify, measure, and operationalize transportation equity is urgently needed as planners and decision-makers are undecided on how to distribute existing transportation resources and simultaneously invest in future transportation infrastructure *fairly*.

The application of equity within the realm of transportation planning has proliferated in the academic literature but concrete adoption of the concepts in planning practice has been minimal [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].
Academic literature has approached equity through measures such as accessibility and activity participation [@paezMeasuringAccessibilityPositive2012; @allenPlanningTransportSocial2020], affordability [@isalouTestingHousingTransportation2014], environmental impacts and safety [@guoSystematicOverviewTransportation2020], and health [@fransenIdentifyingPublicTransport2015; @smithSystematicLiteratureReview2017], and also conceptualizations such as frameworks for equity and social exclusion [@lucasProvidingTransportSocial2006; @lucasTransportPovertyIts2016], and distributive justice theories [@behbahaniDesigningRoadNetwork2019; @lewisExploringEquityHow2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], among others.
However, a gap in the literature is the absence of a comprehensive review that examines how equity is conceptualizated, theorized and operationalized through measures and standards.
These inquiries are vital to planning agencies and decision-makers, and as such, this review aims to definitively benchmark the knowledge in the literature from this perspective.

Though aspects of transport equity in metropolitan planning documents such as accessibility [@boisjolyHowGetThere2017] or mode-specific equity [@doranPursuitCyclingEquity2021] have been reviewed, no review to date captures how equity has been theorized and translated into standards from the perspective of equity dimensions.
Our work contributes in this direction, reviewing which equity dimensions, standards, and conceptualizations have been applied in the academic literature and identifying where gaps exist in the literature.
Ultimately, this review aims to collate the academic knowledge and present it in an operationalizable manner to both academics and decision-makers in hopes to catalyze the uptake in equitable planning practices and guide future research programmes which develop definitions, measures, and recommendations for evaluating transportation equity.

This paper is structured as follows.
In Section 2, we set the stage by defining what do we mean by equity dimensions, conceptualizations and types of standards used in the literature search and throughout the paper.
In Section 3, we outline the methods used in the scoping review conducted.
In Section 4, we summarize the findings of the literature.
In Section 5, we discuss the methods connects to standards, conceptualizations and dimensions.
In Section 6, we summarize the findings and link them to future transport planning agendas in addition to visioning ways the findings can be used to build *just* transportation planning processes.

@@

Transportation equity literature exists within cross-cutting categories.
We borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.
This categorization is outlined in @jaggarPhilosophicalChallengesGlobal2009 to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental. 
Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?". This is discussed as a next step in planning for *justice* transportation literature in Section 6.

Furthermore, it is important to distinguish between equity and justice.
Our focus is on equity, which is nested within and can be interpreted as the application of justice theories.
For instance, *mobility justice* seeks to understand how broader power inequities inform the governance and control of movement are reproduced by mobility systems [@shellerMobilityJusticePolitics2018]; in understanding what is *just* and *unjust,* equity is the application of processes to move towards justice.
What is justice must be decided at a community level, a goal of the equity processes, and cannot be captured in a single academic text.

@@

-   *When* and *Where*: includes the article's publication year, the geographic context (i.e., urban, suburban and rural areas, countries) and location.
-   *Who*: includes the type of population group, mode type, activity/opportunity type, and transport network status.
-   *How*: includes the measure used to assess equity dimensions as supported by a **equity conceptualization** and **equity standard**. **Equity conceptualizations** are some type of motivating equity principle, philosophy or theory. For example: theoretical and conceptual frameworks (e.g., transport-related social exclusion, transport disadvantage, transport poverty, horizontal equity, vertical equity) and equity principles (e.g., Utilitarism, Capabilities Approach, Sufficientarism). **Equity standards** are thresholds that when operationalized can be used to define when an aspect of the transportation system is in/equitable. They can be quantitative thresholds, qualitative descriptions, and/or mixed-method thresholds. Examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), distributional assessment (e.g., composite indicator, range, maximum distance or gap, etc.), dis/satisfaction with travel, etc.

>>>>

> "_To be wealthy and honored in an unjust society is a disgrace._"
> — Confucius, The Annalects

The fiery wake of a rocket could be seen ascending, piercing the blue skies of West Texas. It was mid-morning on July 20, 2021, and on board was a small group of four passengers that included Jeff Bezos, the founder of Amazon and then the world's richest person [@harwoodBezos2021]. The mission that morning was among the first ever crewed private suborbital flights, and the adventure (described as "intense" by a member of the crew) lasted a total of 10 minutes and 10 seconds [@harwoodBezos2021]. In addition to intense, the undertaking was expensive: a seat for the flight had previously been auctioned for no less than $28 million USD [@griffinPerson2021]. Before even 10:00 am (EDT) Bezos was back from his excursion, and right on time to declare that this was the "[b]est day ever". To reporters covering the event he said "I want to thank every Amazon employee and every Amazon customer, because you guys paid for all of this" [@johnsonRoadToSpace2021].

Meanwhile, firmly grounded on planet Earth, the employees that Bezos thanked for his suborbital escapade were struggling with some very mundane problems of their own, none of them as lofty as conflicting schedules that prevented them from flying in rockets [@griffinPerson2021]. According to reports, people employed directly or indirectly by Amazon for warehouse or delivery work had, for years, been treated to "inhumane" conditions [@fungEuropean2018; @scottAmazon2019; @greeneAlgorithm2021], and subjected to surveillance on the job, crushing demands for productivity quotas that led to injury, facing little or no job security, and more dealt with as disposable inputs to Amazon's targets and consumers' desires [@tungAmazon2020; @reeseTeamsters2022; @middletonRobot2023].

Coverage of the July 20 launch by the mainstream media was in many cases uncritical. "We're going to build a road to space so that our kids and their kids can build a future" Bezos declared, before adding "...we need to do that to solve the problems here on Earth" [@johnsonRoadToSpace2021]. Few reporters saw it fit to ask what problems the billionaire planned to solve on Earth, or what kind of future Bezos was trying to build, and for whose children. In other words, the billionaire was not confronted with the question of what did his trip do _for whom_ and _to whom_. A less dispassionate observer might have been excused for wondering (possibly aloud) about the basic _fairness_ of a man amassing a nigh unimaginable fortune that allowed him to build and fly his own spaceship, while masses of his employees were treated as throwaway cogs in the vast apparatus of his logistics empire.

The question of _fairness_, however, is not a simple one, and many of us would have been stumped to explain in a clear and precise way, beyond a disquieting intuition, just why the above picture was disturbing. _Justice_, is a political ideal based on the principle that individuals should be treated in a fair and equitable manner [@gosslingUrban2016, p. 2], giving and receiving whatever they are _due_ [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2]. The political [and contested; @vanoutriveHow2019] nature of the concept presents challenges that are only narrowly amenable to scientific inquiry. For starters, the notion that people are "due" something depends on the values of a society, as embodied in its systems and institutions [@karnerFrom2020]. Values,in turn, are not subject to natural laws, but rather are the result of intersubjective realities, which is to say illusions whose legitimacy derives from a collective will to believe. For example, justice would likely mean something very different in a democratic society, if compared to a society where most owed their all—perhaps including their own time or bodily autonomy—to some collective illusion (e.g., the state, a monarch). As well, the meaning of "justice" would likely differ in yet another society where very few owned most, and most owned very little due to a different illusion (e.g., that wealth equates merit). In these two hypothetical cases, elucidating the meaning of "fair" would in all certainty be beyond the dreams of most, since fair would be what the state, the monarch, or the extremely wealthy said it was. In contrast, in democratic societies^[Paraphrasing, democracy is the worst of all collective illusions, except for all others.] individuals rights are valued above the whims of the few. In such a setting, the task of defining a "just" distribution of the burdens and benefits of things—from income, to roads, to space travel—quickly becomes muddled, encumbered even, by the necessity to pay attention to a multitude of voices, not all of them equally loud.

The rocket that took Bezos to the edge of space is a somewhat rare example of a transportation technology, a tool of space-time convergence. By enabling movement at very high speeds rockets might—one day, at some indeterminate point in the future—prove essential to the expansion of the human species beyond our home planet. But in the present moment, the benefits of a private suborbital flight are for a few (e.g., the joy of movement, the sense of adventure, the awe of seeing Earth from space), whereas the burdens (e.g., the use of non-renewable resources, the climate-altering emissions) affect us all, and not evenly at that. In a plutocratic society, those with most can argue that this distribution of burdens and benefits is fair since both benefits and burdens are earned. The fact that the likes of Bezos can travel to space is proof that the likes of Bezos deserve those trips. In a democratic society, the members of the collective might actually agree that large rewards (e.g., space flight) must be offered to highly qualified individuals (e.g., Bezos) to entice them to take important responsibilities (e.g., founding and leading Amazon). The members of the collective might think this a fair state of affairs, having come to the conclusion of their own accord or after being persuaded by billionaires. But then again, the members of the collective might decide otherwise: the _values_ of those in charge of defining what is "fair" matter.

Multiple national and cross-national studies suggest that people in many societies have some tolerance for inequality [@kiatpongsanHow2014]: indeed, it appears that _some_ stratification, as suggested by @davisSome1945, is thought to fulfill a valuable function. However, extreme inequality is often frowned upon, and can lead to social dysfunction and other ills [@acemogluWhy2000; @taydasCan2012; @duIncome2019; @houleInequality2022]. But the perceptions of what is "fair" are neither universal or static. Instead, they are malleable, and can be affected by the existence of opportunities for social mobility [@shariffIncome2016; @artigeWhat2019], by exposure to inequality [@schroderIncome2017; @garciaChanging2023], by learned helplessness [@kimSocioeconomic2022], and even by how information about inequality is communicated to the public [@walkerPeople2021]. It follows that, in general terms, there are at least three different manners of thinking about inequality: 1) in a _positive_ (or descriptive) manner, as the current state of the distribution of benefits and burdens of things; 2) also descriptively, as the _perceptions_ about the distribution of those benefits and burdens; and 3) normatively (or prescriptively), as the desired or ideal state of the same. Clearly, the three may coincide (e.g, if the perceived levels of inequality matched actual inequality and also how much of it the public desired). However, they do not necessarily have to, and in many cases will differ from one another. Measuring inequality as it is, in the most objective way possible is an essential task to decide whether there is a need to develop inequality-related policy to increase fairness. In turn, measuring the perceptions of inequality, in the most accurate way possible, may be important to achieve sufficient public buy-in in order to enhance the chances that a given policy will succeed.

Transportation systems are a class of essential technologies that facilitate or impede social inclusion and activity participation by the public [@churchTransport2000; @lucasTransport2001; @socialExclusion2003; @cassSocial2005; @casasSocial2007; @prestonAccessibility2007; @paezMobility2009]. As such, they increasingly are of interest from the perspective of equity, and a lively and rapidly growing literature has emerged in response [see _inter alia_ @martensTransport2016; @diciommoTransport2017; @guoSystematicOverviewTransportation2020; @karnerFrom2020; @vecchioTransportEquityLatin2020; @pereiraTransportationEquity2021; @vanweeEvaluating2021; @zhangLiterature2021; @desjardinsExaminingEquityAccessibility2022; @karnerAdvancesPitfallsMeasuring2023]. A cynical rationale for this interest could be that keeping track of actual and perceived inequalities can serve _at the very least_ as a gauge of social discontent [as Chilean authorities discovered to their sorrow when unrest erupted over the increase of metro fares in 2019; @BBCNews2019; @diazpabonInequality2021]. More optimistically, in democratic systems, there could be a genuine interest in being ready to respond to popular demands. Several challenges arise when approaching this endeavor, caused by the notorious complexity of transportation systems: simultaneously, they move people, goods, and information. As well, emerging technologies and service models can swiftly change the balance of benefits and burdens among a population [@guoSystematicOverviewTransportation2020]. Unlike income, which is as unidimensional as it gets, the benefits and burdens of transportation systems are diffuse over space and time. For example, transportation systems engineered to offer higher mobility for people _somewhere_, can simultaneously cut others off essential opportunities _elsewhere_ , as @rajeEngineering2004 poignantly illustred with examples of infrastructure in the UK. Furthermore, the shades of policies past can continue to haunt a region and even the planet for decades or longer, as shown by the legacy of displacement and decay caused by urban highways all across the US [@archerWhite2020] and the time horizon for the impacts of climate change to be fully felt [@markolfTransportation2019].

But complexity is no reason to shirk the task.

The objective of this review is to investigate the state of knowledge in terms of defining and operationalizing "fairness" in the transportation domain. Much research has been devoted to the issues of _measuring _ equity in transportation, including @ramjerdiEquity2006, @delboscUsing2011, @welchMeasure2013, @martensMeasuring2019, and @pritchardNew2022, among many others. Further, there are multiple sources that discuss the conceptual and philosophical foundations of equity and fairness in transportation [e.g., @martensTransport2016; @pereiraDistributiveJusticeEquity2017a; @vanoutriveHow2019]. Finally, previous reviews of planning documents have investigated equity from narrowly scoped perspectives, such as accessibility [@boisjolyHowGetThere2017] or a particular mode of transportation [e.g., cycling; @doranPursuitCyclingEquity2021]. These inquiries are valuable to planning agencies and decision-makers, and the present review will tread similar, but not completely overlapping ground since, in our estimation, there is a gap in the literature in terms of understanding how standards for equity are developed and implemented in the transportation domain. In contrast, we do know that adoption of equity concepts in planning practice has lagged developments in academic work [@pereiraTransportationEquity2021; @boisjolyHowGetThere2017; @doranPursuitCyclingEquity2021; @linovskiEquityTransportationPlanning2020; @litmanEvaluatingTransportationEquity2022].

To illustrate this gap, we note how @beilerExploring2016, in their exploration of transport equity, cite the following strategies identified by the US DOT to address justice [p. 287]:

- Reduce adverse human health and environmental effects on minority and low-income populations.
- Include all potentially affected communities in the transportation decision-making process.
- Ensure that minority and low-income populations receive equitable benefits.

While commendable, the strategies are sufficiently vague that they can be implemented in a myriad ways, either to comply with the spirit of justice, or instead performatively to deceive it [@mcculloughPerformative2023]. Some relevant questions include: how much should the adverse effects be reduced? To zero? Or to some tolerable level of adversity? What should that level be? What are the criteria for deciding that a community is "potentially affected"? What benefits should be distributed, and should those be based on simple population weights? Or, contrariwise, should more deprived individuals be eligible for a larger share of the benefits?

These questions boil down to the development and use of _standards_ for transportation justice. A definition of "standard" is as "something set up and established by authority as a rule for the measure of quantity, weight, extent, value, or quality"^[https://www.merriam-webster.com/dictionary/standard]. How much pollution is allowed to be generated, and where, depends on who is affected, as well as on how much health is valued and by whom. For example, firms may or may not adopt lower standards for the emission of pollutants in poorer areas, or poor people may end up being relegated to areas that adopt lower standards [@gouldsonFirms2006]. The result is the same: pollution tends to be worse were poorer people are [@delucaCode2012].

Supporting the creation of (more) equitable transportation systems involves understanding the production, distribution, and management of transportation benefits and costs; how they are distributed; and what values are implemented (and by whom) in the form of standards [@pereiraDistributiveJusticeEquity2017a; @shellerMobilityJusticePolitics2018; @pereiraTransportationEquity2021]. Thus, for this review we engage the literature with the following questions in mind: 1) what is our current understanding of the things that transportation systems do, for whom, and to whom; 2) what does the literature say about the distribution of the benefits and costs of transportation systems; 3) and what values are embodied in normative statements about said distribution. Ultimately, this review aims to collate the academic knowledge and present it in a manner useful to support both academics and decision-makers in the development and implementation of standards for equity in transportation planning and policy, some of which are sorely missing, even in those places where legislation explicitly calls for justice^[[As recently as -@martensFair2021, Martens and Golub note that Federal directives related to Title VI in the US "do not provide guidelines that can help agencies develop explicit standards to assess the distribution of accessibility benefits from projects or plans."]].

<!--
This paper is structured as follows.
In Section 2, we set the stage by defining what do we mean by equity dimensions, conceptualizations and types of standards used in the literature search and throughout the paper.
In Section 3, we outline the methods used in the scoping review conducted.
In Section 4, we summarize the findings of the literature.
In Section 5, we discuss the methods connects to standards, conceptualizations and dimensions.
In Section 6, we summarize the findings and link them to future transport planning agendas in addition to visioning ways the findings can be used to build *just* transportation planning processes.
-->

@@

> "_I should like to be able to love my country and still love justice_."
> — Albert Camus

> "_Never forget that justice is what love looks like in public_."
> ― Cornel West

<!-- From the original intro

Benefits and burdens can include, but are not limited to, accessibility, mobility, affordability, health and environmental issues that impact individuals, marginalized members of a community, and the community as a whole [@lucasMeasuringTransportEquity2019]. Academic literature has approached equity through measures such as accessibility and activity participation [@paezMeasuringAccessibilityPositive2012; @allenPlanningTransportSocial2020], affordability [@isalouTestingHousingTransportation2014], environmental impacts and safety [@guoSystematicOverviewTransportation2020], and health [@fransenIdentifyingPublicTransport2015; @smithSystematicLiteratureReview2017], and also conceptualizations such as frameworks for equity and social exclusion [@lucasProvidingTransportSocial2006; @lucasTransportPovertyIts2016], and distributive justice theories [@behbahaniDesigningRoadNetwork2019; @lewisExploringEquityHow2021; @pereiraDistributiveJusticeEquity2017a; @vecchioAccessibilityCapabilitiesApproach2021], among others.

-->

## What is justice?

In the paragraphs above we used the terms "justice", "fairness", and "equity" in a relatively loose fashion. This was done purposefuly. As we noted, people often have a strong intution of what is "fair", "just", and "equitable". These conceptions may or may not match those of the authorities who set the standards of fairness. But in democratic societies, the authority of political leaders, bureaucrats, planners, and all those charged with the business of governing, derives from the will of the people. It is therefore important to make the implicit explicit, and clearly spell out those intuitions of "justice", "fairness", and "equity" to make them conceptually clear. Only this way can the participants of a democratic society be effective arbiters of what is just - and perhaps stand on firmer ground to experience love of their society.

Let us begin by stating that justice is an end goal, that is, a desirable state of affairs that we strive to achieve.

Justice is attained when people "give and receive whatever they are due" [@jaggarPhilosophicalChallengesGlobal2009, pp. 1-2], and it ceases to exist when there are persons or groups that are denied "access to the opportunities they need to lead a meaningful and dignified life" [@karnerTransportationEquityTransportation2020, p. 440]

Several forms of justice can be distinguished, including [see @jaggarPhilosophicalChallengesGlobal2009; @karnerTransportationEquityTransportation2020]:

**Retributive justice**. It is concerned with the retribution due to people who do wrongs.

**Reparative justice**. This form of justice relates to the proper way to correct or rectify past wrongs even if the wrongdoers can no longer be meted retribution.

**Procedural justice**. The main concern of this form of justice is to ensure that the views, opinions, and preferences of all stakeholders are properly accounted for when decisions that affect their lives are made.

**Distributive justice**. Perhaps the most commonly studied form of justice [see @jaggarPhilosophicalChallengesGlobal2009, p. 2; @pereiraDistributiveJusticeEquity2017a], and its main concern is the way the benefits and burdens of the tangible and intangible products of society.

It might be argued that all of the above are particular forms of distributive justice. Retributive justice, for example, is often concerned with the distribution of the benefits and burdens of being a member of society: the way it is usually achieved is by distributing intangibles like "freedom" (e.g., of movement, of association) as benefits, and/or claiming tangibles like money as burdens (e.g., fines). Reparative justice often distributes current benefits and burdens to redress past wrongs, for example by asking those who have benefited from past wrongs, even if unwittingly, to shoulder a bigger fiscal burden in order to cover programs that mete benefits to those who are harmed by past wrongs. Procedural justice is the distribution of the benefits (e.g., the right to voice an opinion) and burdens (e.g., the effort required to develop an educated opinion) of the processes that lead to decisions with collective consequences.

We can then speak of the purposes of distributive justice: to mete out retribution _fairly_, to repair past harms, and to ensure that procedures offer _equitable_ opportunities for engagement. Equity and fairness are the instruments of justice, the tools by which society advances towards the end goal of justice.

<!-- Furthermore, it is important to distinguish between equity and justice. -->

The term "equity" tends to encompass various tools to understand the distribution of benefits and burdens of things among a population. In the transportation domain, the term is somewhat loaded because it is perceived as stemming from the authority of the state, and meant to assist with decisions about regulating and financing transportation spending [@karnerTransportationEquityTransportation2020]. Here, we are in agreement with @karnerTransportationEquityTransportation2020 that equity analysis should not be seen as an end in and of itself, but rather as a means to gather information about actual and perceived inequality. In this respect, the analytical tradition of equity, at least in transportation planning, means that the relevant models become embedded in the "political ecology of the estimated truth" [@kingModels1993]: in principle their assumptions and scope must be open and transparent, but they are also vulnerable to misuse.

Fairness, in contrast to equity, is more complicated to define. The concept does not have the same history of development as an analytical tool, and can be interpreted in numerous, and possibly discordant ways. That this is the case is convincingly demonstrated by @martensFair2021 in their study of the application of Title VI of the Civil Rights Act of 1964 in accessibility planning practice in the US. Title VI explicitly talks about the distribution of benefits derived from Federal funding: "[n]o person in the United States shall, on the ground of race, color, or national origin, be excluded from participation in, be denied the benefits of, or be subjected to discrimination under any program or activity receiving Federal financial assistance." However, as @martensFair2021 show, there are several ways to comply with regulations while achieving different outcomes: from a very mild form of fairness that simply avoids intentional discrimination, to strategically overdeliver the benefits to compensate for past discrimination (reparative justice). What justice did fairness serve in each case? It depends on what was the reason for seeking justice in the first place! Our reading of @martensFair2021 is that fairness is a yardstick that is best used _a priori_ than _a posteriori_, for doing so risks rationalizing the outcomes rather than

The focus of this report is on standards of fairness that, combined with the use of equity analysis, can help understand how better to move towards just outcomes in transportation planning.

<!--- EDITED UP TO HERE --->

## A framework for questions of justice

For the analysis we are inspired by the framing of @jaggarPhilosophicalChallengesGlobal2009 for philosophical questions of justice.

borrow a "Where?", "When?", "Who?", "What?", and "How?" framing to categorize aspects of how equity is assessed.

<!-- The texture of these questions becomes more crisp when talking about transportation, which is inherently about space and time, that is to say about "where" and "when". Conventionally, theories of justice have been aspatial, despite an early interest of geographers on matters of justice [@]-->

This categorization is outlined in  to describe substantive sets of answers to questions of distributive justice (and further discussed in the context of justice within the critical human geography discipline [@przybylinskiWhereJusticeGeography2022]).
For questions of *justice*, convincing answers to these questions require a rationale: a "Why?" [@jaggarPhilosophicalChallengesGlobal2009].
For the purpose of this review focused on *equity*, the focus is on studies that substantively answer "Where?", "When?", "Who?", "What?", and "How?"; the answer to "Why?" may be incidental.

## Transportation and justice

To understand the distributional implications of transportation systems, it is essential that we are clear about what they do. In other words, what do transportation systems _produce_?

At their most fundamental, transportation systems are space-time convergence tools, tools that improve the rate at which time is traded for space. They usually do this by increasing the speed of movement: sidewalks facilitate walking, traffic lights facilitate the ordered flow of vehicular traffic, and a launching pad makes it possible for a rocket to takeoff. With complex interlocking parts (sidewalk, road, traffic sign, parking regulations), transportation systems produce _mobility_, the potential for movement. Combined with land use systems (the spatial distribution of opportunities on the landscape), they produce _accessibility_, the potential to reach destinations. The realization of those potentials happens through travel. But, as the adage goes, travel is derived demand, which seems to hold for most (even if not all) travel.

Transportation equity literature exists within cross-cutting categories.

Though critical, answers to "Why?" are not overwhelming common in the transportation equity academic literature that also address  "Where?", "When?", "Who?", "What?", and "How?". This is discussed as a next step in planning for *justice* transportation literature in Section 6.

@@

-   *When* and *Where*: includes the article's publication year, the geographic context (i.e., urban, suburban and rural areas, countries) and location.
-   *Who*: includes the type of population group, mode type, activity/opportunity type, and transport network status.

<!-- Is it appropriate to talk about a "population group"? Think about intersectionality theory.

These are some things that people _are_:

- Gender identity
- Ableness
- Ethnicity
- Age
- Caste
- Low income (in societies with low social mobility)
- Membership in a family

-->

-   *How*: includes the measure used to assess equity dimensions as supported by a **equity conceptualization** and **equity standard**. **Equity conceptualizations** are some type of motivating equity principle, philosophy or theory. For example: theoretical and conceptual frameworks (e.g., transport-related social exclusion, transport disadvantage, transport poverty, horizontal equity, vertical equity) and equity principles (e.g., Utilitarism, Capabilities Approach, Sufficientarism). **Equity standards** are thresholds that when operationalized can be used to define when an aspect of the transportation system is in/equitable. They can be quantitative thresholds, qualitative descriptions, and/or mixed-method thresholds. Examples include: maximum travel distance/cost/time to or from key destinations, levels of maximum exposure to externalities (i.e., noise or air pollution), distributional assessment (e.g., composite indicator, range, maximum distance or gap, etc.), dis/satisfaction with travel, etc.
